Fusing Filters with Integer Linear Programming

### 2014-05-13 2014betterfusionforfilters/section/03-SizeInference.tex
wrap figure[H] around code examples and ILP formulation so they don't split over pages?

The existential quantifier appears in size schemes when the array produced by a filter or similar operator appears in the result. For example:
\begin{alltt}
   filterLeft \(:\sb{s}\,\forall\,k\sb{1}.\,\exists\,k\sb{2}.\,(xs\,:\,k\sb{1})\;\to\;(ys\sb{1}\,:\,k\sb{1},\,ys\sb{2}\,:\,k\sb{2})\)
   filterLeft xs
     = let ys1 = map (+ 1)   xs
           ys2 = filter even xs
       in (ys1, ys2)
\end{alltt}

@@

Finally, note that size schemes form but one aspect of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda we could write something like:
\begin{alltt}
filterLeft : \(\forall\,k\sb{1}:\,\)Nat\(.\,\exists\,k\sb{2}:\,\)Nat\(.\) 
  Array \(k\sb{1}\) Float \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
\end{alltt}

However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process. In our situation, we need automatic type generalization, but for a first order language only.

@@

% -----------------------------------------------------------------------------
\subsection{Rigid Sizes}
When the environment of our size constraints contains rigid variables (indicate by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme. Consider the @filterLeft@ function from S~\ref{s:SizeTypes}
\begin{code}
      filterLeft xs
        = let ys1 = map (+ 1)   xs
              ys2 = filter even xs
          in (ys1, ys2)
\end{code}
The size constraints for this function, already in solved form, are as follows.
$$
\begin{array}{ll}
       & xs : k_{xs},~ ys_1 : k_{ys1},~ \exists k_1,~ ys_2 : k_{ys2},~ k_2
\\
\vdash &          k_{ys_1} = k_1
        ~\wedge~  k_{ys_2} = k_2
        ~\wedge~  k_{xs}   = k_2
\end{array}
$$

@@

When producing size schemes, we do not permit the arguments of an operator graph to have existentially quantified sizes. This restriction is necessary to reject programs that we cannot statically guarantee will be well sized. For example:
\begin{code}
    bad1 xs  = let flt   = filter p xs
                   ys    = map2   f flt xs
               in  ys
\end{code}

The above function filters its input array, and then applies @map2@ to the filtered version, as well as the original array. As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true. The size constraints are as follows:
$$
\begin{array}{ll}
       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_2
\\
\vdash &          k_{flt}  = k_1
        ~\wedge~  k_{flt}  = k_2
        ~\wedge~  k_{xs}   = k_2
        ~\wedge~  k_{ys}   = k_2
\end{array}
$$

@@

\begin{tabbing}
MMMMM \= M \= MMMMMMMMMM \= MM \= \kill
$\iiter_{\Gamma,C}$  
        \>$:$\> $bind \rightarrow T$ 
\\[1ex]
$\iiter_{\Gamma,C}$
        \> $|$  \> $(z~ = @fold@~ f~xs)$     
                \> $=$ \> $C(\Gamma(xs))$ 
\\
        \> $|$  \> $(ys = @map@_n~f~\overline{xs})$
                \> $=$ \> $C(\Gamma(ys))$ 
\\
        \> $|$  \> $(ys = @filter@~f~xs)$    
                \> $=$ \> $C(\Gamma(xs))$ 
\\
        \> $|$  \> $(ys = @generate@~s~f)$  
                \> $=$ \> $C(\Gamma(ys))$ 
\\
        \> $|$  \> $(ys = @gather@~is~xs)$    
                \> $=$ \> $C(\Gamma(is))$ 
\\
        \> $|$  \> $(ys = @cross@~as~bs)$     
                \> $=$ \> $C(\Gamma(as)) \times C(\Gamma(bs))$ 
\\
        \> $|$  \> $(ys = @external@~\overline{xs})$  
                \> $=$ \> $\bot$ 
\\
\end{tabbing}

@@

Looking back at the @normalize@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
%The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
%\gabi{gts is no combinator, so can't be a transducer
The iteration size of @sum2@ is $k_{gts}$, and the filter combinator which produces @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
\begin{code}
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ...
\end{code}

@@

\begin{tabbing}
MMMM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
$trans$  \>$::$\> $binds \rightarrow name \rightarrow \{name\}$ \\
$trans(bs,o)$    \\
            \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $trans'(bs,n)$ \\
            \> $|$ \> otherwise             \>          \> $=$ \> $trans'(bs,o)$ \\
\\
$trans'(bs,o)$    \\
            \> $|$ \> $o = @fold@~f~n$      \> $\in bs$ \> $=$ \> $\emptyset$ \\
            \> $|$ \> $o = @map@_n~f~ns$    \> $\in bs$ \> $=$ \> $\bigcup_{x \in ns} trans(bs, x)$ \\
            \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $\{o\}$       \\
            \> $|$ \> $o = @generate@~s~f$  \> $\in bs$ \> $=$ \> $\emptyset$ \\
            \> $|$ \> $o = @gather@~i~d$    \> $\in bs$ \> $=$ \> $trans(bs,i)$ \\
            \> $|$ \> $o = @cross@~a~b$     \> $\in bs$ \> $=$ \> $\emptyset$ \\
            \> $|$ \> $o = @external@~ins$  \> $\in bs$ \> $=$ \> $\emptyset$ \\
\end{tabbing}

To determine whether two combinators of different iteration sizes may be fused together, we first find parent or ancestor transducers of the same size, if they exist:
\begin{tabbing}
MMMM \= M \= MMMMMMM \= M \= \kill
$parents$ \> $::$ \> $name \to name \to \{name \times name\}$ \\
$parents(a,b)$ \\
        \> $|$ \> $\tau(a) == \tau(b)$ \> $=$    \> $\{(a, b)\}$ \\
        \> $|$ \> otherwise            \> $=$    \> $\{ parents(a', b) ~|~ a' \in trans(a) \} $      \\
        \>     \>                      \> $\cup$ \> $\{ parents(a, b') ~|~ b' \in trans(b) \} $  \\
\end{tabbing}

>>>>

The existential quantifier appears in size schemes when the array produced by a filter or similar operator appears in the result. For example:
\begin{figure}[H]
\begin{alltt}
 filterLeft \(:\sb{s}\,\forall\,k\sb{1}.\,\exists\,k\sb{2}.\,(xs\,:\,k\sb{1})\;\to\;(ys\sb{1}\,:\,k\sb{1},\,ys\sb{2}\,:\,k\sb{2})\)
 filterLeft xs
   = let ys1 = map (+ 1)   xs
         ys2 = filter even xs
     in (ys1, ys2)
\end{alltt}
\end{figure}

@@

Finally, note that size schemes form but one aspect of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda we could write something like:
\begin{figure}[H]
\begin{alltt}
 filterLeft : \(\forall\,k\sb{1}:\,\)Nat\(.\,\exists\,k\sb{2}:\,\)Nat\(.\) 
   Array \(k\sb{1}\) Float \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
\end{alltt}
\end{figure}
However, the type inference systems for fully higher order dependently typed languages typically require quantified types to be provided by the user, and do not perform the type generalization process. In our situation, we need automatic type generalization, but for a first order language only.

@@

% -----------------------------------------------------------------------------
\subsection{Rigid Sizes}
When the environment of our size constraints contains rigid variables (indicate by $\exists k$), we introduce existential quantifiers instead of universal quantifiers into the size scheme. Consider the @filterLeft@ function from S~\ref{s:SizeTypes}
\begin{figure}[H]
\begin{code}
      filterLeft xs
        = let ys1 = map (+ 1)   xs
              ys2 = filter even xs
          in (ys1, ys2)
\end{code}
\end{figure}
The size constraints for this function, already in solved form, are as follows.
$$
\begin{array}{ll}
       & xs : k_{xs},~ ys_1 : k_{ys1},~ \exists k_1,~ ys_2 : k_{ys2},~ k_2
\\
\vdash &          k_{ys_1} = k_1
        ~\wedge~  k_{ys_2} = k_2
        ~\wedge~  k_{xs}   = k_2
\end{array}
$$

@@

When producing size schemes, we do not permit the arguments of an operator graph to have existentially quantified sizes. This restriction is necessary to reject programs that we cannot statically guarantee will be well sized. For example:
\begin{figure}[H]
\begin{code}
    bad1 xs  = let flt   = filter p xs
                   ys    = map2   f flt xs
               in  ys
\end{code}
\end{figure}
The above function filters its input array, and then applies @map2@ to the filtered version, as well as the original array. As the @map2@ operators requires both of its arguments to have the same size, @bad1@ would only be valid when the predicate @p@ is always true. The size constraints are as follows:
$$
\begin{array}{ll}
       & xs : k_{xs},~ flt : k_{flt},~ \exists k_1,~ ys : k_{ys},~ k_2
\\
\vdash &          k_{flt}  = k_1
        ~\wedge~  k_{flt}  = k_2
        ~\wedge~  k_{xs}   = k_2
        ~\wedge~  k_{ys}   = k_2
\end{array}
$$

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM \= M \= MMMMMMMMMM \= MM \= \kill
$\iiter_{\Gamma,C}$  
        \>$:$\> $bind \rightarrow T$ 
\\[1ex]
$\iiter_{\Gamma,C}$
        \> $|$  \> $(z~ = @fold@~ f~xs)$     
                \> $=$ \> $C(\Gamma(xs))$ 
\\
        \> $|$  \> $(ys = @map@_n~f~\overline{xs})$
                \> $=$ \> $C(\Gamma(ys))$ 
\\
        \> $|$  \> $(ys = @filter@~f~xs)$    
                \> $=$ \> $C(\Gamma(xs))$ 
\\
        \> $|$  \> $(ys = @generate@~s~f)$  
                \> $=$ \> $C(\Gamma(ys))$ 
\\
        \> $|$  \> $(ys = @gather@~is~xs)$    
                \> $=$ \> $C(\Gamma(is))$ 
\\
        \> $|$  \> $(ys = @cross@~as~bs)$     
                \> $=$ \> $C(\Gamma(as)) \times C(\Gamma(bs))$ 
\\
        \> $|$  \> $(ys = @external@~\overline{xs})$  
                \> $=$ \> $\bot$ 
\\
\end{tabbing}
\end{figure}

@@

Looking back at the @normalize@ example, the iteration size of @gts@ and @sum1@ are both $k_{xs}$.
%The iteration size of @sum2@ is $k_{gts}$, and @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
%\gabi{gts is no combinator, so can't be a transducer
The iteration size of @sum2@ is $k_{gts}$, and the filter combinator which produces @gts@ is a transducer from $k_{xs}$ to $k_{gts}$. 
Despite it not being known whether $k_{gts} = k_{xs}$, the three nodes @gts@, @sum1@ and @sum2@ can all be fused together.
The two nodes @sum2@ and @gts@ can be fused together, since the output size of @gts@ is the iteration size of @sum2@.
Similarly, @sum1@ and @gts@ can be fused together, as they have the same iteration size.
\begin{figure}[H]
\begin{code}
 normalize2 xs
  = let sum1 = fold   (+)  0   xs
        gts  = filter (> 0)    xs
        sum2 = fold   (+)  0   gts
        ...
\end{code}
\end{figure}

@@

\begin{figure}[H]
\begin{tabbing}
MMMM \= MM \= MMMMMMMMM \= MMMM \= MM \= \kill
$trans$  \>$::$\> $binds \rightarrow name \rightarrow \{name\}$ \\
$trans(bs,o)$    \\
            \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $trans'(bs,n)$ \\
            \> $|$ \> otherwise             \>          \> $=$ \> $trans'(bs,o)$ \\
\\
$trans'(bs,o)$    \\
            \> $|$ \> $o = @fold@~f~n$      \> $\in bs$ \> $=$ \> $\emptyset$ \\
            \> $|$ \> $o = @map@_n~f~ns$    \> $\in bs$ \> $=$ \> $\bigcup_{x \in ns} trans(bs, x)$ \\
            \> $|$ \> $o = @filter@~f~n$    \> $\in bs$ \> $=$ \> $\{o\}$       \\
            \> $|$ \> $o = @generate@~s~f$  \> $\in bs$ \> $=$ \> $\emptyset$ \\
            \> $|$ \> $o = @gather@~i~d$    \> $\in bs$ \> $=$ \> $trans(bs,i)$ \\
            \> $|$ \> $o = @cross@~a~b$     \> $\in bs$ \> $=$ \> $\emptyset$ \\
            \> $|$ \> $o = @external@~ins$  \> $\in bs$ \> $=$ \> $\emptyset$ \\
\end{tabbing}
\end{figure}

To determine whether two combinators of different iteration sizes may be fused together, we first find parent or ancestor transducers of the same size, if they exist:
\begin{figure}[H]
\begin{tabbing}
MMMM \= M \= MMMMMMM \= M \= \kill
$parents$ \> $::$ \> $name \to name \to \{name \times name\}$ \\
$parents(a,b)$ \\
        \> $|$ \> $\tau(a) == \tau(b)$ \> $=$    \> $\{(a, b)\}$ \\
        \> $|$ \> otherwise            \> $=$    \> $\{ parents(a', b) ~|~ a' \in trans(a) \} $      \\
        \>     \>                      \> $\cup$ \> $\{ parents(a, b') ~|~ b' \in trans(b) \} $  \\
\end{tabbing}
\end{figure}

### 2014-05-13 2014betterfusionforfilters/section/04-ILP.tex
wrap figure[H] around code examples and ILP formulation so they don't split over pages?

\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A1) (B2)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}

@@

\begin{tabbing}
MMMM        \= MM   \= \kill
$nodes$     \> @::@ \> $program \to V$          \\
$edges$     \> @::@ \> $program \to E$          \\
$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
\\
MMMMM       \= \kill
$nodes(bs)$ \> $= \{(name(b), \tau_b) | b \in bs\}$       \\
\end{tabbing}
Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
Creating edges between these nodes is simply when one binding references an earlier one. The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.

\begin{tabbing}
MMMMM       \= \kill
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(b)$    \\
\\
MMMM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
                              \> $=$    \> $\{(out,data, fusion~preventing) \}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
                              \> $=$    \> $\{inedge(bs, out, a) \}$       \\
                              \> $\cup$ \> $\{(out, b, fusion~preventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
                              \> $=$    \> $\{(outs,i, fusion~preventing) | i \in ins \}$ \\
\\
$inedge(bs,to,from)$ \\
                     \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $(\ldots,from,\ldots = @external@ \ldots) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $otherwise$                      \\
                     \> $=$ \> $(to, from, fusible)$            \\
\end{tabbing}

@@

As an example of a cyclic clustering, consider:
\begin{code}
cycle us
 = let xs  = map (+1) us          (C1)
       sum = fold xs              (C2)
       zs  = map (+sum) xs        (C1)
\end{code}
There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
If @xs@ and @zs@ were clustered together in @C1@ and @sum@ were in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.

@@

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (an edge from $i$ to $j$)            \\
           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (no edge from $i$ to $j$)            \\
           \> \ldots
\end{tabbing}
As per Megiddo\cite{megiddo1998optimal}, we look at every pair of nodes $i$ and $j$.

@@

\paragraph{Fusion-preventing edges}
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (fusion-preventing edges from $i$ to $j$)      \\
           \> \ldots
\end{tabbing}
As per Megiddo\cite{megiddo1998optimal}, for every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur. This, together with the precedence constraints above, has the effect of enforcing $\pi_i < \pi_j$.

@@

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \emptyset$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \{i',j'\}$) \\
           \> \ldots
\end{tabbing}
If two nodes have different iteration sizes, they may not necessarily be fused.
However, if they have parent transducers of the same size, they may be fused if both their parents are fused together, and each is fused with its respective parent.

@@

As a degenerate example, consider fusing an operation on filtered data with its generating filter:
\begin{code}
sum1 = fold (+) 0  us
gts  = filter (>0) us
sum2 = fold (+) 0  gts
\end{code}
Here, $parents(gts,sum2) = \{gts, gts\}$.
This generates spurious, but still valid, constraints that $gts$ must be fused with $gts$ and $gts$ must be fused with $sum2$, in order for $gts$ and $sum2$ to be fused together.
While these constraints are unnecessary in this case, they are harmless.

@@

\paragraph{Array contraction}
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i)            \\
           \> \ldots \\
\end{tabbing}
As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
An array is fully contracted if all its output edges are fused with it. Thus, $c_i=0$ only if $\forall j | (i,j) \in E. x_{ij} = 0$. By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays. \ben{Say what array contraction is, and why we want this constraint.}

@@

\begin{tabbing}
MMMMM   \= MMMM \= M \= \kill
Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
           \> \> \> (memory traffic and loop overhead)         \\
           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
           \> \> \> (removing intermediate arrays)         \\
Subject to \> \ldots                                \\
Where      \>                                       \\
           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        \\
           \\
           \> $N = |V|$
\end{tabbing}

@@

\begin{tabbing}
MMMMMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
$split$     \> @::@ \> $V \times E \to \{ \{name\} \}$      \\
MMMM        \= M    \= \kill
$split(vs,es)$ \\
    \> $=$  \> $\{clusterable(v,vs,es) | v \in vs\}$      \\
$clusterable(v,vs,es)$  \\
    \>$=$       \>$\{v' | v' \in vs$                                \\
    \>$\wedge$  \>$\forall p. p = path(v,v') \vee p = path(v',v)$   \\
    \>$\wedge$  \>$fusion~preventing \not\in p\}$\\
\end{tabbing}

@@

\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (no edge from i to j, but there is a fusion benefit)            \\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (an edge from i to j)            \\
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \>             (a \emph{fusion-preventing} edge from i to j)            \\
\\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i with no fusion-preventing outputs)      \\
\\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
\end{tabbing}

>>>>

\begin{figure}[H]
\begin{code}
 normalizeInc :: Array Int -> Array Int
 normalizeInc xs
  = let incs = map  (+1)    us      (A1) (B2)
        sum  = fold (+) 0   us      (A1) (B1)
        norm = map  (/ sum) incs    (A2) (B2)
    in  norm
\end{code}
\end{figure}

@@

\begin{figure}[H]
\begin{tabbing}
MMMM        \= MM   \= \kill
$nodes$     \> @::@ \> $program \to V$          \\
$edges$     \> @::@ \> $program \to E$          \\
$edge$      \> @::@ \> $\{bind\} \times bind \to E$\\
\\
MMMMM       \= \kill
$nodes(bs)$ \> $= \{(name(b), \tau_b) | b \in bs\}$       \\
\end{tabbing}
\end{figure}
Each node may simply be the name of its output binding (or bindings, in the case of @external@); as we require names to only be bound once, this is assured to be unique.
Creating edges between these nodes is simply when one binding references an earlier one. The only complication is designating edges as \emph{fusible} or \emph{fusion-preventing}.

\begin{figure}[H]
\begin{tabbing}
MMMMM       \= \kill
$edges(bs)$ \> $= \bigcup_{b \in bs}edge(b)$    \\
\\
MMMM             \= M \= \kill
$edge(bs, out = @fold@~f~in)$ \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @map@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @filter@~f~in)$  \\
                              \> $=$    \> $\{inedge(bs,out,s) | s \in f\}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, in) \}$       \\
$edge(bs, out = @gather@~data~indices)$  \\
                              \> $=$    \> $\{(out,data, fusion~preventing) \}$ \\
                              \> $\cup$ \> $\{inedge(bs, out, indices) \}$       \\
$edge(bs, out = @cross@~a~b)$            \\
                              \> $=$    \> $\{inedge(bs, out, a) \}$       \\
                              \> $\cup$ \> $\{(out, b, fusion~preventing) \}$ \\
$edge(bs, outs = @external@~ins)$  \\
                              \> $=$    \> $\{(outs,i, fusion~preventing) | i \in ins \}$ \\
\\
$inedge(bs,to,from)$ \\
                     \> $|$ \> $(from = @fold@~f~s) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $(\ldots,from,\ldots = @external@ \ldots) \in bs$     \\
                     \> $=$ \> $(to, from, fusion~preventing)$  \\
                     \> $|$ \> $otherwise$                      \\
                     \> $=$ \> $(to, from, fusible)$            \\
\end{tabbing}
\end{figure}

@@

As an example of a cyclic clustering, consider:
\begin{figure}[H]
\begin{code}
cycle us
 = let xs  = map (+1) us          (C1)
       sum = fold xs              (C2)
       zs  = map (+sum) xs        (C1)
\end{code}
\end{figure}
There is no fusion-preventing edge directly between @xs@ and @zs@, but there is a fusion-preventing edge between @sum@ and @zs@.
If @xs@ and @zs@ were clustered together in @C1@ and @sum@ were in cluster @C2@, there would be a dependency cycle between @C1@ and @C2@, and neither could be executed before the other.

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (an edge from $i$ to $j$)            \\
           \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (no edge from $i$ to $j$)            \\
           \> \ldots
\end{tabbing}
\end{figure}
As per Megiddo\cite{megiddo1998optimal}, we look at every pair of nodes $i$ and $j$.

@@

\paragraph{Fusion-preventing edges}
\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (fusion-preventing edges from $i$ to $j$)      \\
           \> \ldots
\end{tabbing}
\end{figure}
As per Megiddo\cite{megiddo1998optimal}, for every fusion-preventing edge, we add the constraint $x_{ij} = 1$, so that no fusion can occur. This, together with the precedence constraints above, has the effect of enforcing $\pi_i < \pi_j$.

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $=$   \> $1$             \>       \>            \\
           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \emptyset$)  \\
           \\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> $x_{i'j'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau(i) \not= \tau(j) \wedge parents(i,j) = \{i',j'\}$) \\
           \> \ldots
\end{tabbing}
\end{figure}
If two nodes have different iteration sizes, they may not necessarily be fused.
However, if they have parent transducers of the same size, they may be fused if both their parents are fused together, and each is fused with its respective parent.

@@

As a degenerate example, consider fusing an operation on filtered data with its generating filter:
\begin{figure}[H]
\begin{code}
 sum1 = fold (+) 0  us
 gts  = filter (>0) us
 sum2 = fold (+) 0  gts
\end{code}
\end{figure}
Here, $parents(gts,sum2) = \{gts, gts\}$.
This generates spurious, but still valid, constraints that $gts$ must be fused with $gts$ and $gts$ must be fused with $sum2$, in order for $gts$ and $sum2$ to be fused together.
While these constraints are unnecessary in this case, they are harmless.

@@

\paragraph{Array contraction}
\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> \ldots \\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i)            \\
           \> \ldots \\
\end{tabbing}
\end{figure}
As per Darte's work on optimising for array contraction\cite{darte2002contraction}, we define a variable $c_i$ for each array.
An array is fully contracted if all its output edges are fused with it. Thus, $c_i=0$ only if $\forall j | (i,j) \in E. x_{ij} = 0$. By minimising $c_i$ in the objective function, we favour solutions that reduce the number of required intermediate arrays. \ben{Say what array contraction is, and why we want this constraint.}

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMMM \= M \= \kill
Minimise   \>     \> $\Sigma_{(i,j) \in E} W_{ij} x_{ij}$   \\
           \> \> \> (memory traffic and loop overhead)         \\
           \> $+$ \> $\Sigma_{i \in V} N c_i$  \\
           \> \> \> (removing intermediate arrays)         \\
Subject to \> \ldots                                \\
Where      \>                                       \\
           \> $W_{ij} = N^2$ \> $~|$ \> $(i,j) \in E $         \\
           \> \> \> (fusing $i$ and $j$ will reduce memory traffic)         \\
           \> $W_{ij} = N^2$ \> $~|$ \> $\exists k. (k,i) \in E \wedge (k,j) \in E $     \\
           \> \> \> ($i$ and $j$ share an input array)                                         \\
           \> $W_{ij} = 1$   \> $~|$ \> $@otherwise@$                                                  \\
           \> \> \> (the only benefit is loop overhead)                                        \\
           \\
           \> $N = |V|$
\end{tabbing}
\end{figure}

@@

\begin{figure}[H]
\begin{tabbing}
MMMMMM      \= MM   \=  MMMMMMMMMMMM    \=  \kill
$split$     \> @::@ \> $V \times E \to \{ \{name\} \}$      \\
MMMM        \= M    \= \kill
$split(vs,es)$ \\
    \> $=$  \> $\{clusterable(v,vs,es) | v \in vs\}$      \\
$clusterable(v,vs,es)$  \\
    \>$=$       \>$\{v' | v' \in vs$                                \\
    \>$\wedge$  \>$\forall p. p = path(v,v') \vee p = path(v',v)$   \\
    \>$\wedge$  \>$fusion~preventing \not\in p\}$\\
\end{tabbing}
\end{figure}

@@

\begin{figure}[H]
\begin{tabbing}
MMMMM   \= MMM \= M \= MMM \= M \= MMM \= \kill
Minimise   \> \ldots \\
Subject to \> $-N x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (no edge from i to j, but there is a fusion benefit)            \\
           \>    $x_{ij}$ \> $\le$ \> $\pi_j - \pi_i$ \> $\le$ \> $N x_{ij}$ \\
           \>             (an edge from i to j)            \\
           \>             \>       \> $\pi_i < \pi_j$ \>       \>            \\
           \>             (a \emph{fusion-preventing} edge from i to j)            \\
\\
           \> $x_{ij}$    \> $\le$ \> $c_i$           \>       \>            \\
           \> (for all edges from i with no fusion-preventing outputs)      \\
\\
           \> $x_{i'j}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_i=\{i'\}$) \\
           \> $x_{ij'}$   \> $\le$ \> $x_{ij}$        \>       \>            \\
           \> (if $\tau_i \not= \tau_j \wedge gen_j=\{j'\}$) \\
\end{tabbing}
\end{figure}
